---
title: "Statistical Analysis: Feedback Processing"
output: html_document
date: "2026-08-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


# Preparation
In this section, we load the packages required for the subsequent analyses.

```{r}
# Data manipulation
library(tidyverse)
library(tidyr)
library(dplyr)

# Descriptive statistics
library(psych)

# Visualization
library(ggplot2)
library(corrplot)

# MIxed models
library(lme4)
library(lmerTest)

# Estimated marginal means
library (emmeans)
```


```{r}
# Set fixed colours for our feedback conditions

feedback_colour <- "cornflowerblue"
control_colour <- "coral"

feedback_colours <- c(
  "Control" = control_colour,
  "Feedback" = feedback_colour
)
```



# Experimental Data

## Loading the Dataset

The experimental dataset containing reaction times and accuracy values is loaded from a CSV file.

```{r}
descr_df <- read_csv("Exp_longlist_merged_2026_07_02.csv", show_col_types = FALSE)
# descr_df
```


## Descriptive Statistics

Two main measures are analysed:

- **Accuracy** (whether the response was correct)
- **Reaction time (RT)** in milliseconds


### Initial Inspection of the Data
Before conducting statistical analyses, the dataset is filtered to include only experimental trials containing real words. This removes practice items and nonword trials.

```{r}
# Remove all practice items, remove all nonword trials
pre_filtered_df <- subset(descr_df, Item_type == "exp" & Condition == "word")
pre_filtered_df
```

### Participant-Level Summary Statistics

To identify potential outliers or problematic participants, we compute the mean accuracy and mean reaction time for each participant.

This allows us to detect:

- participants with extremely low accuracy (indicating misunderstanding of the task)
- participants with unusually slow responses

A preliminary plot is used to visualize potential outliers.

```{r participant_outliers, cache=TRUE}
# Overall overview by participant
describeBy(pre_filtered_df, group = pre_filtered_df$Participant)

# Summary: mean accuracy and RT by participant
means_by_participant <- pre_filtered_df %>%
  group_by(Participant) %>%
  summarise(
    mean_RT = mean(RT, na.rm = TRUE),
    mean_acc = mean(Accuracy, na.rm = TRUE)
  )
means_by_participant

# Plot
plot_outliers <- function(y_axis, filename, colour){
  p <- ggplot(data = means_by_participant, aes(x = "", y = {{ y_axis }})) +
  geom_boxplot(fill = colour, color = "black", alpha = 0.6) +
  labs(
    x = "",
    y = "Mean by Participant"
  ) +
  theme_minimal()

ggsave(
  filename,
  plot = last_plot(),
  width = 7,
  height = 5
)

p
}

plot_outliers(mean_RT, "average_RT_unfiltered.png", "grey70")
plot_outliers(mean_acc, "average_acc_unfiltered.png", "grey70")
```

No participants are excluded because the participant-level descriptive statistics do not indicate problematic performance.

## Feedback coding
The feedback variable is coded as a factor with control as the reference category. This means that the model coefficients for Feedback represent the difference between the Feedback and Control conditions.

```{r feedback_coding, cache=TRUE}
filtered_df <- pre_filtered_df %>%
  mutate(
    Feedback = factor(
      Feedback,
      levels = c(0, 1),
      labels = c("Control", "Feedback")
    )
  )

filtered_df

```




### Reaction Time Filtering

To remove implausible reaction times, a cutoff is applied.

Only trials with reaction times between **200 ms and 2000 ms** are retained.

This removes extremely fast responses (likely accidental key presses) and extremely slow responses that may reflect distractions or interruptions.

```{r reaction_time_filtering, cache=TRUE}
# Visualize the distribution of reaction times to see where to place the cutoff
p <- ggplot(data = filtered_df, aes(x = "", y = RT)) +
  geom_boxplot(fill = "grey70", color = "black", alpha = 0.6) +
  labs(
    x = "",
    y = "Reaction Time"
  ) +
  theme_minimal()

#ggsave(
#  "RT_distribution.png",
#  plot = last_plot(),
#  width = 7,
#  height = 5
#)

p



# Apply cutoff
post_filtered_df <- filtered_df %>%
  filter(RT >= 200 & RT <= 2000)



# Plot again to visualize the new distribution
p <- ggplot(data = post_filtered_df, aes(x = "", y = RT)) +
  geom_boxplot(fill = "grey70", color = "black", alpha = 0.6) +
  labs(
    x = "",
    y = "Reaction Time"
  ) +
  theme_minimal()

#ggsave(
#  "RT_outlier.png",
#  plot = last_plot(),
#  width = 7,
#  height = 5
#)

p

```


### Mean Accuracy and Reaction Time by Feedback Condition
Next, mean accuracy and reaction time are computed for each feedback category:

- 1: feedback
- 0: control


```{r mean_acc_RT_by_feedback, cache=TRUE}
# Mean accuracy and reaction times by feedback condition
means_by_feedback <- post_filtered_df %>%
  group_by(Feedback) %>%
  summarise(
    mean_RT = mean(RT, na.rm = TRUE),
    mean_acc = mean(Accuracy, na.rm = TRUE)
  )
means_by_feedback

# Plot mean accuracies (barplots)
p <- ggplot(
  means_by_feedback,
  aes(
    x = Feedback,
    y = mean_acc,
    fill = Feedback
  )
) +
  geom_col(
    width = 0.7
  ) +
  scale_fill_manual(
    values = feedback_colours
  ) +
  scale_x_discrete(
    limits = c("Feedback", "Control")
  ) +
  labs(
    x = "Feedback Condition",
    y = "Mean Accuracy",
    fill = "Feedback"
  ) +
  theme_minimal()

ggsave(
  "average_acc_per_feedback_barplot.png",
  plot = p,
  width = 7,
  height = 5
)

p


# Plot mean RTs (boxplots)
p <- ggplot(
  post_filtered_df,
  aes(
    x = Feedback,
    y = RT,
    fill = Feedback
  )
) +
  geom_boxplot(
    width = 0.7
  ) +
  scale_fill_manual(
    values = feedback_colours
  ) +
  scale_x_discrete(
    limits = c("Feedback", "Control")
  ) +
  labs(
    x = "Feedback Condition",
    y = "Reaction Time (ms)",
    fill = "Feedback"
  ) +
  theme_minimal()

ggsave(
  "RT_per_feedback_boxplot.png",
  plot = p,
  width = 7,
  height = 5
)

p
```

# Inferential Statistics
To formally test the research questions, mixed-effects models are used.

Mixed-effects models are appropriate for experimental data because they
allow us to account for variability across both:

- participants
- stimuli


## Data Loading and Preparation

- Data loading: The full experimental dataset containing reaction times and accuracy values as well as participant data is loaded from a CSV file.
- Filtering: everything that was removed from the above dataset needs to be removed here as well (practice items, nonwords, certain participants)
- Variable coding: Before fitting the models, variables are converted to appropriate types (e.g., numeric or factor). The ordering of factor levels is also specified for interpretability.

```{r}
prepped_df <- post_filtered_df %>%
  mutate(
    Accuracy = as.numeric(Accuracy),
    RT = as.numeric(RT),

    Participant = factor(Participant),
    Stimulus = factor(Stimulus),

    Feedback = factor(
      Feedback,
      levels = c("Control", "Feedback")
    ),

    # Block_cat = factor(Block),           # tried to have the individual Blocks as a factorial variable but it does not work as well as having it continuous

    Block_cont = as.numeric(Block),
    anxiety_score = as.numeric(anxiety_score),
    openness_to_experience = as.numeric(openness_to_experience),
    conscientiousness = as.numeric(conscientiousness),
    extraversion = as.numeric(extraversion),
    agreeableness = as.numeric(agreeableness),
    neuroticism = as.numeric(neuroticism)
  )

prepped_df
```

### Standardizing Continuous Predictors

Continuous predictors are standardized (z-transformed). This improves model convergence and ensures that predictors are on comparable scales.
The continuous individual-difference variables are standardized before entering them into the models. This makes the coefficients easier to compare and can improve model estimation.


```{r}
prepped_df <- prepped_df %>%
  mutate(
    anxiety_z = as.numeric(scale(anxiety_score)),
    openness_z = as.numeric(scale(openness_to_experience)),
    conscientiousness_z = as.numeric(scale(conscientiousness)),
    extraversion_z = as.numeric(scale(extraversion)),
    agreeableness_z = as.numeric(scale(agreeableness)),
    neuroticism_z = as.numeric(scale(neuroticism)),
  )
prepped_df
```


## Accuracy Analysis

Accuracy is analysed using a generalized linear mixed-effects model with a binomial distribution.

The primary question is whether feedback affects accuracy. Block is included to examine whether performance changes over the course of the experiment, and the interactions between Feedback and the individual-difference variables test whether personality or anxiety moderates the feedback effect.

The model includes:

Fixed effects:

- feedback
- block (time effect of being given feedback)
- each of the Big Five scores, plus the anxiety score


Random effects:

- random intercepts for participants
- random intercepts for stimuli


```{r ACC_mixed_model, cache=TRUE}
acc_model <- glmer(
  Accuracy ~
    Feedback *
    Block_cont +
    anxiety_z +
    Feedback:anxiety_z +
    openness_z +
    Feedback:openness_z +
    conscientiousness_z +
    Feedback:conscientiousness_z +
    extraversion_z +
    Feedback:extraversion_z +
    agreeableness_z +
    Feedback:agreeableness_z +
    neuroticism_z +
    Feedback:neuroticism_z +
    (1 | Participant) +
    (1 | Stimulus),
  family = binomial,
  data = prepped_df,
  control = glmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 1e5)
  )
)

summary(acc_model)

```

```{r}
# Plot the model

block_seq <- seq(
  min(prepped_df$Block_cont, na.rm = TRUE),
  max(prepped_df$Block_cont, na.rm = TRUE),
  by = 0.1
)

# Estimated marginal means
# Personality variables are held at their mean (0)
emm <- emmeans(
  acc_model,
  ~ Feedback | Block_cont,
  at = list(
    Block_cont = block_seq,
    anxiety_z = 0,
    openness_z = 0,
    conscientiousness_z = 0,
    extraversion_z = 0,
    agreeableness_z = 0,
    neuroticism_z = 0
  ),
  type = "response"
)

# Convert to dataframe
emm_df <- as.data.frame(emm)

# Plot
p <- ggplot(
  emm_df,
  aes(
    x = Block_cont,
    y = prob,
    color = Feedback,
    fill = Feedback
  )
) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(
    aes(
      ymin = asymp.LCL,
      ymax = asymp.UCL
    ),
    alpha = 0.2,
    color = NA
  ) +
  scale_color_manual(
    values = feedback_colours
  ) +
  scale_fill_manual(
    values = feedback_colours
  ) +
  labs(
    x = "Block",
    y = "Predicted Accuracy",
    color = "Feedback",
    fill = "Feedback"
  ) +
  scale_x_continuous(
    breaks = 1:6
  ) +
  scale_y_continuous(
    limits = c(0, 1)
  ) +
  theme_minimal(base_size = 14)

ggsave(
  "acc_model_feedback_block.png",
  plot = p,
  width = 7,
  height = 5
)

p
```





## Reaction Time Analysis

Reaction times are analysed using a **linear mixed-effects model**.

The primary question is whether feedback affects accuracy. Block is included to examine whether performance changes over the course of the experiment, and the interactions between Feedback and the individual-difference variables test whether personality or anxiety moderates the feedback effect.

Only trials with correct responses are included, as incorrect trials may reflect different cognitive processes.

The model includes:

Fixed effects:

- feedback
- block (time effect of being given feedback)
- each of the Big Five scores, plus the anxiety score


Random effects:

- random intercepts for participants
- random intercepts for stimuli


```{r RT_data}
RT_df <- prepped_df %>%
  filter(Accuracy == 1)

RT_df

```

The model uses the same predictors as the accuracy model.


```{r RT_mixed_model, cache=TRUE}
RT_model <- lmer(
  RT ~
    Feedback *
    Block_cont +
    anxiety_z +
    Feedback:anxiety_z +
    openness_z +
    Feedback:openness_z +
    conscientiousness_z +
    Feedback:conscientiousness_z +
    extraversion_z +
    Feedback:extraversion_z +
    agreeableness_z +
    Feedback:agreeableness_z +
    neuroticism_z +
    Feedback:neuroticism_z +
    (1 | Participant) +
    (1 | Stimulus),
  data = RT_df
)

summary(RT_model)

```

```{r}
# Plot the model
# Block as a continuous predictor

block_seq <- seq(
  min(RT_df$Block_cont, na.rm = TRUE),
  max(RT_df$Block_cont, na.rm = TRUE),
  by = 0.1
)

# Estimated marginal means
# Personality variables are held at their mean (0)
emm <- emmeans(
  RT_model,
  ~ Feedback | Block_cont,
  at = list(
    Block_cont = block_seq,
    anxiety_z = 0,
    openness_z = 0,
    conscientiousness_z = 0,
    extraversion_z = 0,
    agreeableness_z = 0,
    neuroticism_z = 0
  )
)

# Convert to dataframe
emm_df <- as.data.frame(emm)

# Plot
p <- ggplot(
  emm_df,
  aes(
    x = Block_cont,
    y = emmean,
    color = Feedback,
    fill = Feedback
  )
) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(
    aes(
      ymin = lower.CL,
      ymax = upper.CL
    ),
    alpha = 0.2,
    color = NA
  ) +
  scale_color_manual(
    values = feedback_colours
  ) +
  scale_fill_manual(
    values = feedback_colours
  ) +
  labs(
    x = "Block",
    y = "Predicted RT (ms)",
    color = "Feedback",
    fill = "Feedback"
  ) +
  scale_x_continuous(
    breaks = 1:6
  ) +
  theme_minimal(base_size = 14)

ggsave(
  "RT_model_feedback_block.png",
  plot = p,
  width = 7,
  height = 5
)

p
```



# Interpretation

The results of the mixed-effects models are interpreted with respect to
the research questions:

1. Does summary accuracy feedback affect lexical decision performance, as reflected in reaction times and accuracy?
2. Do Big Five personality traits and foreign language anxiety moderate the effect of feedback on performance?
3. Does the effect of feedback change over the course of the experiment?

The answers to these questions are based on the inferential statistical results reported above.


